Easy as Pi: A Network Coding Raspberry Pi Testbed

### 2016-06-02 schemes.tex
Theory done for now

In this section, we present a description of the considered coding schemes
measured with our the Raspberry Pi testbed. We introduce a definition
for the primitive coding operations, e.g. encoding, decoding and recoding
(where applies) for each coding scheme. Later, we address particular schemes
which are obtained by modifying the basic coding operations that provide
better processing speeds, particularly relevant for the Raspberry Pi.

@@

\subsubsection{Method 1: Fixing the Coding Density}
A way to control the amount of non-zero coding coefficients, is to set a
fixed ratio of non-zero coefficients in the encoding vector of size $g$. We
refer to this fixed ratio as the average coding density $d$. Thus, for any
coded packet $C_i$ with coding coefficients $v_{ij},\ j \in [1,g]$, its
coding density is defined as follows:
%
\begin{align}
\label{eq:coding_density}
d = \frac{\sum_{j=1}^{g} f(v_{ij})}{g},\ \
    f(v_{ij}) =
    \begin{cases}
        0 & \text{for } v_{ij} = 0 \\
        1 & \text{for } v_{ij} \neq 0
    \end{cases}
\end{align}
%
From the density definition in \eqref{eq:coding_density}, it can be
observed that $0 \leq d \leq 1$. As $g$ increased, we obtain more
granularity in the density. Notice that the special case $d = 0$ has no
practical purpose since it implies just zero padded data. So, a practical
range for the density excludes the zero case, $0 < d \leq 1$. Moreover,
it can be also observed that the case of $d = 1$ reduces to the \ac{RLNC}
coding scheme described in Section \ref{ssec:RLNC}.

To achieve a desired average coding density in the coefficients, for each
of them, we utilize a set of Bernoulli random variables all with parameter
$d$ as it success probability, i.e. $\mathbb{B}_j \sim Bernoulli(d)\ ,\ \forall
j \in [1,g]$. In this way, we can represent a coded packet in \ac{SRLNC} as:
%
\begin{align} \label{eq:sparse_coded_packet}
    C_i  &= \bigoplus_{j=1}^{g} \mathbb{B}_j v_{ij} \otimes P_j ,\ v_{ij}
    \neq 0\
    \forall i \in [1,\ldots) ,\ \
    d \in
    \begin{cases}
        {(0,0.5]} & \text{for } q = 2 \\
        {(0,1]} & \text{for } q > 2
    \end{cases}
\end{align}
%
In \eqref{eq:sparse_coded_packet}, we have the requirement for the coding
coefficient to not be zero, since we want to ensure a coding coefficient
is generate for any random trial where $\mathbb{B}_j = 1$. Therefore, in our
implementation of \ac{SRLNC} we exclude the zero element and then pick
uniformly distributed random elements from $GF(q)-\{ 0 \}$. Also, we have
specified a dependency on the field size for practical
density values. In case of employing $GF(2)$, the
maximum plausible density is restricted up to 0.5 since higher values incur
in more frequent \ac{l.d.} coded packets \textbf{Include reference here}.

@@

Provided a desired budget $B$ and the \ac{dof}s of the
decoder, an encoder can estimate the required coding density to not exceed
the required y the budget. ow sparse the coded packets should be to finish a transmission in
time.

In our implementation, it is the encoders responsibility to estimate the coding
density. We use feedback to provide the encoder a better estimate of the
decoders \ac{dof} at pre-defined points. Namely, when the decoder obtains the
following \ac{dof}'s:

\begin{equation}\label{eq:feedback_dofs}
    s(k) = g\cdot \frac{2^k -1}{2^k},  \; \text{for } k=0,1,2,\dots
\end{equation}
where $k$ is the $k$-th feedback packet that is transmitted during a generation.

The encoder considers the time between feedback packets as regions. It then
estimates a fixed coding density for each region when a feedback packet is
received.

Before the coding density can be estimated it is essential that the encoder
has a good estimate of the decoders \ac{dof}, but also the remainder of the
total budget. The feedback scheme \ref{eq:feedback_dofs} roughly splits the
remaining \ac{dof} in halves. This is also the case for the budget that is
split equally among the two new regions. The very last region will be assigned
the full remainder of the total budget.

\subsection{Network Coding Implementation for Multicore Architectures}
\label{sub:implementation-multicore}

The operations needed to encode and decode data are, in general, similar. To
encode packets, the encoder needs to perform the matrix multiplication
$\textbf{C} = \textbf{V} \cdot \textbf{P}$. On the other hand, decoding the
information requires the decoder to find $\textbf{V}^{-1}$ and to perform the
matrix multiplication $\textbf{P} = \textbf{V}^{-1} \cdot \textbf{C}$. In both
cases, a matrix multiplication is needed. Therefore, to make a practical
implementation of network coding, it is valuable to find a way to optimize the
matrix multiplications operations for multicore architectures.

When designing multithreaded algorithms for network coding operations, it is
possible to implement the decoding by combining the matrix inversion and the
matrix multiplication, e.g., performing the Gauss-Jordan algorithm over the
coding matrix $\textbf{V}$ while performing, in parallel, row operations on the
coded data $\textbf{C}$. For example, in~\cite{5061951} and~\cite{4262451} the
parallelization of the row operations are optimized for \ac{GPU} and \ac{SMP}
systems respectively. However, the parallelization of such operations provide
limited speeds up for small block sizes ($\leq 2048$ bytes). The reason is that
to operate in a parallel fashion over the same coded packet $C_i$ requires
strained synchronization.

Therefore, to overcome the constraints of a tight synchronization, it is
preferable to explicitly invert the matrix $\textbf{V}$, and then take advantage
of optimizations for matrix multiplications, both at encoding and decoding time.
With that purpose, the authors implemented in~\cite{wunderlich2015network} an
algorithm that adopts the ideas of efficient \ac{BLAS}
operations~\cite{lawson1979basic} reimplementing them for finite fields
operations. Although there are already certain libraries available such as
~\cite{dumas2008dense} and \cite{dumas2002linbox} that allows the usage of the
highly optimize \ac{BLAS} implementations for some finite fields, they work
converting the GF elements into floating point numbers and back. Even though
that approach is efficient for large matrix sizes, the overhead imposed by the
conversions is not suitable for the the matrix sizes of typical network coding
implementations. The implemented algorithm aims to be cache efficient by
maximizing the number of operations performed over a block of fetched data. This
is done by dividing the matrices in square sub-blocks and operating with them.
As a consequence, this technique exploits the spatial locality of the data, at
least for $\mathcal{O}(n^3)$ algorithms~\cite{golub2012matrix}. The optimal size
of the block is architecture-dependent. The ideal block has a number of operands
that fit into L1 cache, and it is a multiple of the \ac{SIMD} operations size.

The main idea of the implemented algorithm is to represent each one of the sub-
block matrix operations (matrix-matrix multiplication, matrix-triangle matrix
multiplication, triangle-matrix system solving, etc) as a base, \textit{kernel},
operation that can be optimized individually, e.g., using \ac{SIMD} operations.
Each of these kernel operations, at the same time, can be represented as a task
with inputs and outputs in memory. If any algorithm is formulated
conventionally, and the data operations on the sub-blocks are reordered, then it
is possible to resolve data dependencies with a \ac{DAG}. For the algorithm, a
dependency means, for example, that a sub-block of the matrix in memory cannot
be read unless a previous writing operation was performed there. When the
\ac{DAG} is determined, then, a scheduler assigns to the available cores the
execution of the individual kernel operations that have their dependencies
satisfied. The benefit of this method is that the synchronization relies only on
data dependencies, and it does not requires the insertion of artificial
synchronization points. Using this technique, the matrix inversion is performed
using an algorithm based on LU
factorization~\cite{Dongarra:2011:HPM:2132876.2132885}, and the matrix
multiplication is performed by making the various matrix-matrix multiplications
on the sub-blocks.

>>>>

In this section, we present a description of the considered coding schemes
measured with our the Raspberry Pi testbed. We introduce a definition
for the primitive coding operations, e.g. encoding, decoding and recoding
(where applies) for each coding scheme. Later, we address particular schemes
which are obtained by modifying the basic coding operations that provide
better processing speeds, particularly relevant for the Raspberry Pi.
Finally, we include a review of the

@@

\subsubsection{Method 1: Fixing the Coding Density}
A way to control the amount of non-zero coding coefficients, is to set a
fixed ratio of non-zero coefficients in the encoding vector of size $g$. We
refer to this fixed ratio as the average coding density $d$. Thus, for any
coded packet $C_i$ with coding coefficients $v_{ij},\ j \in [1,g]$, its
coding density is defined as follows:
%
\begin{align}
\label{eq:coding_density}
d = \frac{\sum_{j=1}^{g} f(v_{ij})}{g},\ \
    f(v_{ij}) =
    \begin{cases}
        0 &,\ v_{ij} = 0 \\
        1 &,\ v_{ij} \neq 0
    \end{cases}
\end{align}
%
From the density definition in \eqref{eq:coding_density}, it can be
observed that $0 \leq d \leq 1$. As $g$ increased, we obtain more
granularity in the density. Notice that the special case $d = 0$ has no
practical purpose since it implies just zero padded data. So, a practical
range for the density excludes the zero case, $0 < d \leq 1$. Moreover,
it can be also observed that the case of $d = 1$ reduces to the \ac{RLNC}
coding scheme described in Section \ref{ssec:RLNC}.

To achieve a desired average coding density in the coefficients, for each
of them, we utilize a set of Bernoulli random variables all with parameter
$d$ as it success probability, i.e. $\mathbb{B}_j \sim Bernoulli(d)\ ,\ \forall
j \in [1,g]$. In this way, we can represent a coded packet in \ac{SRLNC} as:
%
\begin{align} \label{eq:sparse_coded_packet}
    C_i  &= \bigoplus_{j=1}^{g} \mathbb{B}_j v_{ij} \otimes P_j ,\ v_{ij}
    \neq 0\
    \forall i \in [1,\ldots) ,\ \
    d \in
    \begin{cases}
        {(0,0.5]} & ,\ q = 2 \\
        {(0,1]}   & ,\ q > 2
    \end{cases}
\end{align}
%
In \eqref{eq:sparse_coded_packet}, we have the requirement for the coding
coefficient to not be zero, since we want to ensure a coding coefficient
is generate for any random trial where $\mathbb{B}_j = 1$. Therefore, in our
implementation of \ac{SRLNC} we exclude the zero element and then pick
uniformly distributed random elements from $GF(q)-\{ 0 \}$. Also, we have
specified a dependency on the field size for practical
density values. In case of employing $GF(2)$, the
maximum plausible density is restricted up to 0.5 since higher values incur
in more frequent \ac{l.d.} coded packets \textbf{Include reference here}.

@@

Provided a desired budget $B$ and the \ac{dof} of the decoder, an encoder
can estimate the required coding density for the packets to not exceed the
budget. In our implementation, we use feedback packets to provide the
encoder an estimate of the decoder \ac{dof} at pre-defined points of the
transmission process. The points in the process occur when a decoder
obtains a given amount of \ac{dof}. In our implementation, we define $r(k),\
k \in [1, \ldots)$ as the rank $r$ of the decoder\footnote{The rank of
a decoder is the rank of its coding matrix at a given point in the
transmissions process.} where a decoder has to send the $k$-th feedback in
\eqref{eq:feedback_dofs}.
%
\begin{equation}\label{eq:feedback_dofs}
    r(k) = \floor{g\cdot \left[\frac{2^k -1}{2^k}\right]},\ k \in [1,\ldots)
\end{equation}
%
At the beginning of the transmission process, we assume that the encoder
starts with an initial coding density that has been calculated depending to
the budget. According to \eqref{eq:feedback_dofs}, a decoder sends feedback
once it has obtained (rounded down) $g/2,\ 3g/4,\ 7g/8,\ \ldots$  degrees of
freedom. The total amount of feedback for this scheme will depend on the
generation size. Still, we have an implementation that limits the amount of
feedback to be sent from the decoder in [\textbf{reference}].

The transmissions made between feedback packets are called the density
regions since the same estimated density is used for encoding sparse
packets. Once new feedback is received, an encoder recalculates the coding
density for the new region until the next feedback packet is received.
Before the coding density can be estimated, it is essential that the
encoder has a good estimate of the decoder \ac{dof}, but also the remainder
of the total budget. The feedback scheme based on the rank reports of
\eqref{eq:feedback_dofs} roughly splits the remaining coding regions in
two halves in every. In other words, the first half is the region where
the encoder currently operates. Once an encoder finishes with this region,
it proceeds with second half where again it splits this into to two region
and so on until it finishes on transmitting the whole generation. This is
also the case for the budget that is split equally among the two new
regions. The very last region will be assigned the full remainder of
the total budget and the coding density will not vary.

\subsection{Network Coding Implementation for Multicore Architectures with
the Raspberry Pi}
\label{sub:implementation-multicore}

The arithmetic operations needed to encode and decode data are,
in general, similar. To encode packets, the encoder needs to perform the
matrix multiplication $\textbf{C} = \textbf{V} \textbf{P}$. On the other
hand, decoding the information requires the decoder to find
$\textbf{V}^{-1}$ and to perform the matrix multiplication
$\textbf{P} = \textbf{V}^{-1} \textbf{C}$. In both cases, a matrix
multiplication is needed. Therefore, to make a practical implementation of
network coding, it is valuable to find a way to optimize the matrix
multiplications operations for multicore architectures.

When designing multi-threaded algorithms for network coding operations, it is
possible to implement the decoding by combining the matrix inversion and the
matrix multiplication, e.g. performing the Gauss-Jordan algorithm over the
coding matrix $\textbf{V}$ while performing, in parallel, row operations on the
coded data $\textbf{C}$. For example, in~\cite{5061951} and~\cite{4262451} the
parallelization of the row operations are optimized for \ac{GPU} and \ac{SMP}
systems respectively. However, the parallelization of such operations provide
limited speeds up for small block sizes ($\leq 2048$ bytes). The reason is that
to operate in a parallel fashion over the same coded packet $C_i$ requires
strained synchronization.

Therefore, to overcome the constraints of tight synchronization, a
preferable option is to explicitly invert the matrix $\textbf{V}$, and then
take advantage of optimizations for matrix multiplications, both at
encoding and decoding time. With that purpose, authors
in~\cite{wunderlich2015network} implemented an algorithm that adopts
the ideas of efficient \ac{BLAS}~\cite{lawson1979basic} operations
reimplementing them for finite fields operations. Although there are
libraries such as \cite{dumas2008dense} and \cite{dumas2002linbox} that allow
highly optimized finite field \ac{BLAS} implementations, they work on
converting the \ac{GF} elements into floating point numbers and back.
Even though the approach is efficient for large matrix sizes, the
numerical type conversion overhead is not suitable for matrix sizes of
network coding implementations.

The implemented algorithm in~\cite{wunderlich2015network} aims to be
cache-efficient by maximizing the number of operations performed over a
fetched data block. Here, the matrices are divided in square sub-blocks
where we can operate each of them. As a consequence, this technique
exploits the spatial locality of the data, at least for
$\mathcal{O}(n^3)$ algorithms~\cite{golub2012matrix}. The optimal block
size is architecture-dependent. The ideal block has a number of operands
that fit into the system L1 cache, and it is a multiple of the \ac{SIMD}
operations size.

The idea of the implemented algorithm is to represent each one of the
sub-block matrix operations: matrix-matrix multiplication,
matrix-triangle matrix multiplication, triangle-matrix system solving,
etc; as a base or \textit{kernel} operation that can be optimized
individually using \ac{SIMD} operations. Each kernel operation, at the
same time, can be represented as a task with inputs and outputs in memory.
If an algorithm is formulated conventionally and the data operations on the
sub-blocks are reordered, then it is possible to resolve data
dependencies with a \ac{DAG}. For the algorithm, as an example, a dependency
could be that a sub-block of the matrix in memory cannot be read unless
a previous writing operation was performed there. When the
\ac{DAG} is determined, then a scheduler assigns to the available cores the
execution of the kernel operations that have their dependencies satisfied.
The benefit of this method is that the synchronization relies only on
data dependencies and it does not require the insertion of artificial
synchronization points. Using this technique, the matrix inversion is
performed using an algorithm based on LU factorization
\cite{Dongarra:2011:HPM:2132876.2132885}, and the matrix multiplication is
performed by making the various matrix-matrix multiplications on the
sub-blocks.
